---
title: "FeatureSelection"
author: "Amyre"
date: "2025-03-24"
output: html_document
---

# Clean Data

## Filter

***Read in relative OTU table***

```{r}
rel_data<- read.csv("../otu_rel.csv")
```

***Remove irrelevant rows***

```{r}
rel_data<- rel_data[, 2:(ncol(rel_data)-1) ]
```

***Sum rows***

```{r}
rel_row_sums<- rowSums(rel_data)
```

***Get total***

```{r}
total<- sum(rel_row_sums)
```

***Filter***

Only keep OTUs that make up more than 0.5% of the observed microbes

```{r}
filtered_data <- rel_data[rel_row_sums / total > 0.005, ]
```

## Merge

***Read in relative OTU data***

***Flip table***

In the original table rows are OTUs and columns beetle samples. To match standard OTU format this should be switched.

```{r}
otu_table<- as.data.frame(t(filtered_data))

```

***Adjust row and column names***

```{r}
#columns
col_names <- paste("X.OTU.ID.", colnames(otu_table), sep="")
colnames(otu_table) <- col_names
```

```{r}
#rows
otu_table$BeetleID <- rownames(otu_table)
```

***Read in mapping data***

```{r}
mapping <- read.table("../input/mapping.txt", header = F, sep = "\t")

```

***Get BeetleID and Species information***

```{r}
mapping$BeetleID <- paste("X", mapping$V1, sep="")
```

```{r}
#Extract  species information using regex 
mapping$Species <- sub("^([^ ]+ [^ ]+).*", "\\1", mapping$V4)

```

```{r}
#make new df of important info
keep_mapping<- mapping[, c("BeetleID", "Species")]
```

***Merge otu_table and keep_mapping***

```{r}
merged_data<- merge(otu_table, keep_mapping, by= "BeetleID", all=TRUE)
```

***Remove untreated dung samples***

```{r}
merged_data<- subset(merged_data, Species != "untreated dung")

merged_data$Species<- as.factor(merged_data$Species)
```

These samples are not relevant to our research question.

***Remove BeetleID***

```{r}
merged_data<- merged_data[, 2:ncol(merged_data)]
```

***Check for NAs***

```{r}
sum(is.na(merged_data))
```

***Ensure numbers are numeric***

```{r}
for (col_name in colnames(merged_data)[1:(ncol(merged_data)-1)]) {
  merged_data[[col_name]] <- as.numeric(merged_data[[col_name]])
}
```

## Save

```{r}
write.csv(merged_data, file = "../otu_filtered_merged.csv", row.names = FALSE)
```
